# Rationale and Background

Semaglutide, a glucagon-like peptide-1 receptor agonist (GLP-1 RA), is approved by the US Food and Drug Administration (FDA) for the treatment of type 2 diabetes mellitus. It has been shown to reduce the rate of cardiovascular death, nonfatal myocardial infarction, and nonfatal stroke in patients with type 2 diabetes at high cardiovascular risk.[@marso_semaglutide_2016] Furthermore, results from the recent FLOW trial suggest that semaglutide can reduce clinically important kidney outcomes and death from cardiovascular causes in patients with type 2 diabetes mellitus and chronic kidney disease.[@marso_semaglutide_2016; @perkovic_effects_2024] Given their benefit, the American Diabetes Association’s Standards of Medical Care in Diabetes recommends GLP-1 RA in patients with type 2 diabetes mellitus with atherosclerotic cardiovascular disease and chronic kidney disease.[@american_diabetes_association_professional_practice_committee_9_2021]

Despite the benefits of semaglutide, it has been linked to several adverse ophthalmic consequences previously studied by the OHDSI group.[@cai_semaglutide_2025] Most recently, a retrospective study published in JAMA Ophthalmology linked the exposure of GLP-1 RAs to a two-fold increased risk of neovascular age related macular degeneration (NVAMD) among patients 66 and older.[@shor2025glucagon] The mechanism of this association is unclear, possible theories include the exacerbation of underlying hypoxic mechanisms or increased level of CXCL12, a chemokine that typically rises under hypoxic conditions and promotes angiogenesis and specifically boosts VEGF expression.[@shor2025glucagon]

Conversely, recent studies suggest GLP-1 RA may have anti-inflammatory properties, perhaps through reduced production of pro-inflammatory cytokines, such as IL-6, IL-1beta, TNF-alpha, and CRP, and increased production of anti-inflammatory cytokines, such as IL-10.[@mehdi2023glucagon] There have been recent protective associations between GLP-1 RA use and inflammatory bowel disease,[@bendotti2022anti] and as the GLP-1 receptor is expressed in the eye, there may be a direct protective effect.

Given the common usage of GLP-1 RAs, specifically semaglutide, there exists a need to clarify whether there is an association with NVAMD or uveitis in a larger retrospective cohort study.


# Study Objectives

The purpose of this study is to investigate NVAMD from use of semaglutide for the treatment of type 2 diabetes mellitus.

1. Among patients with type 2 diabetes mellitus, is the risk for NVAMD higher among patients who use semaglutide as compared with a non-GLP-1 RA anti-diabetic medication?
2. Among patients with type 2 diabetes mellitus, is the risk for NVAMD higher among patients who use semaglutide as compared with other GLP-1 RA anti-diabetic medications?

The study also will investigate the incidence of non-infectious uveitis (NIU) from use of semaglutide for the treatment of type 2 diabetes mellitus. 

1. Among patients with type 2 diabetes mellitus, is the risk for NIU lower among patients who use semaglutide as compared with a non-GLP-1 RA anti-diabetic medication?
2. Among patients with type 2 diabetes mellitus, is the risk for NIU lower among patients who use semaglutide as compared with other GLP-1 RA anti-diabetic medications?



# Research Methods

## Study Design/Data Source(s)

This will be a retrospective cohort study employing 2 study designs: 1) an active comparator, new-user cohort design, comparing the risk of NVAMD or NIU in adults with types 2 diabetes mellitus exposed to semaglutide compared with other GLP-1 RA medications and non-GLP-1 RA anti-diabetic medications, and 2) a self-controlled case-series. Other GLP-1 RA medications will include dulaglutide and exenatide. Non-GLP-1 RA comparators will include other second-line treatments for diabetes including empagliflozin (a SGLT2 inhibitor), sitagliptin (a DPP4 inhibitor), and glipizide (a sulfonylurea). These comparators were chosen based on the most commonly occurring medication within those classes from the LEGEND-T2DM study. (https://ohdsi-studies.github.io/LegendT2dm/Protocol)

## Study Population

We will include all adults (≥ 18 years) with type 2 diabetes mellitus in each data source who meet the cohort entry criteria for each exposure cohort (detailed below).

## Exposures {#exposures}
The exposure is receipt of semaglutide (GLP-1 RA), dulaglutide (GLP-1 RA), exenatide (GLP-1 RA), empagliflozin (a SGLT2 inhibitor), sitagliptin (a DPP4 inhibitor), or glipizide (a sulfonylurea) based on the definitions of LEGEND-T2DM. (https://ohdsi-studies.github.io/LegendT2dm/Protocol) For the new-user active comparator cohort design, these will include patients with T2DM withC prior metformin monotherapy who initiate treatment with one of the four medications. Patients must have one year of observation prior to the index date, no prior drug exposure to a comparator second-line or anti-diabetic agents, and at most 30 days of insulin exposure before the drug exposure date.

As a sensitivity analysis, we will examine patients with each of these anti-diabetic medication exposures but without the second-line therapy requirement, meaning that they can be on another anti-diabetic medication at the time of cohort entry, not required to be on prior metformin, but still at most have 30 days of insulin exposure.

## Outcomes {#outcomes}

1. Specific NVAMD
2. Sensitive NVAMD
3. Non-infectious Ocular Inflammation
4. Anterior Uveitis
5. Chorioretinal Inflammation
6. Panuveitis
7. Retinal Vasculitis
8. Non-infectious Ocular Inflammation (excluding intraocular surgery)
9. Anterior Uveitis (excluding intraocular surgery)
10. Chorioretinal Inflammation (excluding intraocular surgery)
11.  Panuveitis (excluding intraocular surgery)
12. Retinal Vasculitis (excluding intraocular surgery)

See human-readable description of the outcomes in the Cohort Definitions section below.


## Covariates

We will deploy the large-scale propensity score method that adjusts for a comprehensive range of baseline characteristic.[@tian2018evaluating]


## Negative Controls

Negative controls will be picked to represent exposure-outcome pairs where no causal effect is expected to exist. We will use the set of negative controls that was used in our prior study of semaglutide and ophthalmic complications.[@protocol_knit_2025]

# Data Analysis Plan

## General

Various standardized analytics available in the OHDSI community will be applied. The Strategus pipeline will be used to call various packages in the HADES library for A) data characterization (A1‑cohort diagnostics, A2‑cohort features, A3‑incidence rates, A4-time-to-event), B) population-level effect estimation (C1–comparative cohort study, C2‑‑self-controlled case-series)[@noauthor_ohdsistrategus_2024; @schuemie_health-analytics_2024] The R package versioning history will be recorded using the renv file and stored in the study Github repository.


## Data Characterization

### Cohort Diagnostics

The R package CohortDiagnostics will be executed for all target, comparator, indication, and outcome cohorts to evaluate measurement error in the phenotype development and evaluation process.

### Cohort Features

The Characterization and FeatureExtraction packages in HADES will be used to identify features of patients in each exposure group who had the outcome and those who did not have the outcome. Features will be extracted from the 365 day observation prior to cohort entry and will include variables such as demographic data (sex, age group, race, ethnicity), prior conditions/drug exposures/procedures/measurements/devices/observations, and risk scores (e.g., Charlson comorbidity index, DCSI, CHADS2VASC score).

### Incidence Rates

The Characterization package in HADES will be used to calculate the incidence rates of the outcome for each exposure cohort and indication cohort. The rate is calculated as the number of outcomes during the time-at-risk divided by the number of total person days.

#### Calculation of time-at-risk

The primary method of calculating time-at-risk will be while the patient is “on-treatment”. This is defined as from the start of cohort entry (or diabetes drug initiation) until the end of the continuous drug exposure, defined as a gap in exposure of >30 days, end of continuous observation period, or study period.

### Time-to-Event

Time to the outcomes of NVAMD will be calculated for each exposure cohort.[@reps_design_2018]

## Population-Level Effect Estimation

### Comparative Cohort Study

The CohortMethod and Cyclops packages in Hades will be used. Large-scale propensity score methods will be used to match the target exposure cohort with the comparator export cohort (e.g., semaglutide vs empagliflozin) using 1:1 propensity score matching.[@tian2018evaluating] Given the concern for increasing usage of semaglutide in recent years, we will also perform a sensitivity analysis with calendar year restriction: Dec2017-Jan2020, Feb2020-June2021, July2021-Dec2024. These dates were chosen based on the date of FDA approval for semaglutide for T2DM, the COVID pandemic, and eventual uptick in semaglutide prescriptions. Cox proportional hazards models will be used to estimate the risk of NVAMD while on treatment using the intent-to-treat design. Negative controls will be used to assess residual bias.

Similar analysis will be performed using NIU as the outcome.


### Self-Controlled Case Series

The SelfControlledCaseSeries and Cyclops packages in Hades will be used to estimate the relative incidence of NVAMD using a self-controlled case-series design. Patients with T2DM and NVAMD diagnosed after the first year in the observation period will be included in the analysis. We will use Poisson regression models to compare the incidence rates of NVAMD while on-treatment with a T2DM medication relative to control time during the observation period. Control time will include periods during the study period when the patient had diagnosed T2DM and excludes the first 365 days of the observation period. Models will have separate adjustments for the pre-exposure time window and potential effects of seasonality and calendar time. Diagnostics will be run to evaluate whether this was an appropriate analysis.

Similar analysis will be performed using NIU as the outcome.


# Study Diagnostics

Residual bias can still remain in observational studies even after PS adjustment is applied to control for measured confounding.[@schuemie_empirical_2018; @schuemie_robust_2016] To address this, we conduct negative control (falsification) outcome experiments for each comparison and outcome, assuming the null hypothesis of no differential effect (i.e., risk ratio (RR) = 1) is true for each negative control outcome. We identified 96 negative controls using a data-driven algorithm that selects OMOP condition concept occurrences with similar prevalence to the outcomes of interest but lacking evidence of association with exposures in published literature, drug product labels, and spontaneous reports.[@voss_accuracy_2017] These were then verified through expert review. The list of negative controls is provided below. From these experiments, we derive an empirical null distribution to account for residual study bias. Using this empirical null, we calibrate each RR estimate, its 95% confidence interval (CI), and the p-value to test for the null hypothesis. We consider an RR significantly different from the null if the calibrated p-value is below 0.05, without adjusting for multiple testing.

To ensure the reliability and generalizability of all comparisons, we evaluate study diagnostics while blinded to the results, and only present estimates that successfully pass these diagnostics.[@schuemie_how_2020; @schuemie_improving_2018] For the primary analysis using the active comparator new-user cohort design, our diagnostics include: (1) preference score distributions between the target and comparator cohorts to evaluate empirical equipoise and population generalizability; (2) cohort balance before and after PS adjustment, defined by the absolute standardized mean differences (SMDs) on extensive patient characteristics for each target-comparator-analysis; (3) negative control calibration plots to assess residual bias, quantified by the Expected Absolute Systematic Error (EASE) derived from the empirical null distribution; and for the primary analysis (Cox PH model), (4) Kaplan-Meier plots to visually examine hazard ratio proportionality assumptions. A study passes diagnostics and contributes to the meta analysis if more than 10% of patients have preference scores between 0.3 and 0.7 on both arms, maximum SMD < 0.1 after PS adjustment, and EASE < 0.25. For the self-controlled case-series analysis, study diagnostics include (1) time trend check that tests for stable background hazards over time periods after PS adjustment, and (2) pre-exposure check to ensure comparative effects in 30 days pre-exposure do not significantly differ from the null to rule out reverse causality; a study passes diagnostics if the p-values for both checks are > 0.05. For both analyses we further evaluate the meta-analytic minimally detectable risk ratio (MDRR) as a proxy of statistical power and only admit a meta-analytic RR estimate if meta-analytic MDRR < 10.[@protocol_knit_references_2025]


## Diagnostic Thresholds for Cohort Method

```{r diagnostic-thresholds-cohort, echo=FALSE, message=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
**Study Diagnostics**; **Threshold**

Covariate balance (Standardized Mean Difference); <0.1
Empirical Equipoise (Preference Score); >0.1
Residual Bias (Expected Absolute Systematic Error (EASE)); <0.25
Meta-analysis Heterogeneity; <0.4
Meta-analysis Minimum Detectable Relative Risk (MDRR); <10
")

tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"),
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

## Diagnostic Thresholds for Self-Controlled Case-Series

```{r diagnostic-thresholds-sccs, echo=FALSE, message=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
**Study Diagnostics**; **Threshold**

Pre-exposure; >0.05
Time Trend; >0.05
Residual Bias (Expected Absolute Systematic Error (EASE)); <0.25
Meta-analysis Minimum Detectable Relative Risk (MDRR); <10
")

tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = c("striped", "hold_position"),
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Strengths and Limitations {#strengths-limitations}

When executed, this will be the largest study examining the association between semaglutide and the risk of NVAMD and NIU. Other strengths of this study include the diversity of the sample (from multiple institutions, multiple countries), and rigorous study design.

Limitations are mainly due to the retrospective nature of the study, which might have residual confounding despite propensity score matching and stratified analyses.


# Protection of Human Subjects

Each participating institution will seek IRB approval for this study as dictated by local governance.

# Plans for Disseminating and Communicating Study Results

The results will be shared and discussed among the study participants and broader OHDSI community during the weekly Tuesday community calls. This work will be presented at conferences (OHDSI Global Symposium and/or ARVO) and published as a manuscript.
